---
title: "Procrustes analyses graphs using ggplot2"
author: "Stephanie Martinez"
date: "22/2/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here we show how to plot the procrustes analysis and consensus shape using "*ggplot2".* Also, we show how to create a .gif and .mp4 showing the transition between shapes across different lakes.

### Packages and data

```{r packages, message=FALSE, warning=FALSE, results='hold'}
library("geomorph")
library("tidyverse")
library("dplyr")
library("data.table")
library("artyfarty")
library("ggpubr")
library("ggplot2")
library("patchwork")

body_epun <- readland.tps("data/body_epun.TPS", specID = "ID", warnmsg =T)
head_epun <- readland.tps("data/head_epun.TPS", specID = "ID", warnmsg =T)
body_ajan <- readland.tps("data/body_ajan.TPS", specID = "ID", warnmsg =T)
head_ajan <- readland.tps("data/head_ajan.TPS", specID = "ID", warnmsg =T)
```

### Procrustes analyses

Now we can generate the procrustes analysis for each species and view

```{r procrustes, message=FALSE, warning=FALSE}
bepun_gpagen <- gpagen(body_epun, Proj=TRUE, print.progress = FALSE)
hepun_gpagen <- gpagen(head_epun, Proj=TRUE, print.progress = FALSE)
bajan_gpagen <- gpagen(body_ajan, Proj=TRUE, print.progress = FALSE)
hajan_gpagen <- gpagen(head_ajan, Proj=TRUE, print.progress = FALSE)
```

This is how the plot looks using the basic **plot** function

```{r}
plot(bepun_gpagen)
plot(hepun_gpagen)
```

So to make this using *ggplot* we need to extract the different dataframes from the gpagen funtion output.

First, we can extract the consensus shape coordinates and the procrustes analysis coordinates to use it in our plots.

```{r consensus}
# Consensus
con_bepun <- as.data.frame(bepun_gpagen$consensus)
con_hepun <- as.data.frame(hepun_gpagen$consensus)
con_bajan <- as.data.frame(bajan_gpagen$consensus)
con_hajan <- as.data.frame(hajan_gpagen$consensus)
# Procrustes coordinates
pro_bepun <- as.data.frame(two.d.array(bepun_gpagen$coords))
pro_hepun <- as.data.frame(two.d.array(hepun_gpagen$coords))
pro_bajan <- as.data.frame(two.d.array(bajan_gpagen$coords))
pro_hajan <- as.data.frame(two.d.array(hajan_gpagen$coords))
```

Second, reshape the coordinates dataframe (showing code only for *E. puntang*)

```{r reshaping}

landm_body<-18
landm_head <- 12

#body
var1 <- as.data.frame(matrix(c("x","y"), landm_body, 2, byrow=T)) %>%   mutate(x=paste(.$V1, rownames(.), sep=""),
         y=paste(.$V2, rownames(.), sep=""))
var2 <- as.vector(t(var1[,-1:-2])) 
names(pro_bepun) <- var2

#head
var3 <- as.data.frame(matrix(c("x","y"), landm_head, 2, byrow=T)) %>%   mutate(x=paste(.$V1, rownames(.), sep=""),
         y=paste(.$V2, rownames(.), sep=""))
var4 <- as.vector(t(var3[,-1:-2])) 
names(pro_hepun) <- var4

```

```{r epun}
#Body
coords_bepun <- pro_bepun %>% 
  mutate(ID=row.names(pro_bepun)) %>%
  pivot_longer(-ID,names_to = "variable",values_to = "value") %>%
  separate(variable, into=c("axis","point"), sep=1, convert = TRUE) %>% 
  pivot_wider(names_from = axis, values_from = value) %>% 
  separate(ID, into = c("lake", "goby","side"), sep = "_", remove = FALSE) %>%
  dplyr::select(-side)
#Head
coords_hepun <- pro_hepun %>% 
  mutate(ID=row.names(pro_hepun)) %>%
  pivot_longer(-ID,names_to = "variable",values_to = "value") %>%
  separate(variable, into=c("axis","point"), sep=1, convert = TRUE) %>% 
  pivot_wider(names_from = axis, values_from = value) %>% 
  separate(ID, into = c("lake", "goby","side"), sep = "_", remove = FALSE) %>%
  dplyr::select(-side)
```

```{r ajan,message=FALSE, warning=FALSE, include=FALSE}
names(pro_bajan) <- var2
names(pro_hajan) <- var4
#Body
coords_bajan <- pro_bajan %>% 
  mutate(ID=row.names(pro_bajan)) %>%
  pivot_longer(-ID,names_to = "variable",values_to = "value") %>%
  separate(variable, into=c("axis","point"), sep=1, convert = TRUE) %>% 
  pivot_wider(names_from = axis, values_from = value) %>% 
  separate(ID, into = c("lake", "goby","side"), sep = "_", remove = FALSE) %>%
  dplyr::select(-side)

#Head
coords_hajan <- pro_hajan %>% 
  mutate(ID=row.names(pro_hajan)) %>%
  pivot_longer(-ID,names_to = "variable",values_to = "value") %>%
  separate(variable, into=c("axis","point"), sep=1, convert = TRUE) %>% 
  pivot_wider(names_from = axis, values_from = value) %>% 
  separate(ID, into = c("lake", "goby","side"), sep = "_", remove = FALSE) %>%
  dplyr::select(-side)
```

### Plotting

Now we can plot using *ggplot2*

```{r plot epun}

plot_bepun <- ggplot() + theme_scientific() + ylim(-0.3,0.3) +
  geom_point(data=coords_bepun,
             aes(x=x,y=y)) +
  geom_point(data=con_bepun,
             aes(x=X,y=Y), size = 3.5, color="deepskyblue") +
  ggtitle("E. puntang - Body view")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank())
plot_hepun <- ggplot() + theme_scientific() + ylim(-0.5,0.5) +
  geom_point(data=coords_hepun,
             aes(x=x,y=y)) +
  geom_point(data=con_hepun,
             aes(x=X,y=Y), size = 3.5, color="darksalmon") +
  ggtitle("E. puntang - Head view")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank())

epun <- plot_bepun + plot_hepun + plot_layout(widths = c(1,1))
epun
ggsave("figures/supplements/procrustes_ggplot_epun.pdf")

```

```{r plot ajan}

plot_bajan <- ggplot() + theme_scientific() + ylim(-0.3,0.3) +
  geom_point(data=coords_bajan,
             aes(x=x,y=y)) +
  geom_point(data=con_bajan,
             aes(x=X,y=Y), size = 3.5, color="deepskyblue") +
  ggtitle("A. janthinopterus - Body view")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank())
plot_hajan <- ggplot() + theme_scientific() + ylim(-0.5,0.5) +
  geom_point(data=coords_hajan,
             aes(x=x,y=y)) +
  geom_point(data=con_hajan,
             aes(x=X,y=Y), size = 3.5, color="darksalmon") +
  ggtitle("A. janthinopterus - Head view")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank())

ajan <- plot_bajan + plot_hajan
ajan
ggsave("figures/supplements/procrustes_ggplot_ajan.pdf")

```

### Animating plots

Now the fun part begins! we can animate and transitions between plots. For example, we can observe how the shape shifts between lakes.

We first need to load the following packages:

```{r animate, message=FALSE, warning=FALSE}
library("gganimate")
library('gifski')
library("png")
library("av") #for video rendering
```

### *Exyrias puntang*

```{r, message=FALSE, warning=FALSE}
#Body
plot_bepun2 <-ggplot(data=coords_bepun,aes(x=x,y=y))+ 
  theme_scientific()+ ylim(-0.3,0.3) +
  geom_point(aes(colour=lake))+ scale_color_brewer(palette="Dark2")+
  ggtitle('E. puntang body - Lake {closest_state}')+
  transition_states(lake,
                    transition_length = 2,
                    state_length = 3.5) +
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),
        legend.position = "bottom")
#Head
plot_hepun2 <-ggplot(data=coords_hepun,aes(x=x,y=y))+ 
  theme_scientific()+ ylim(-0.5,0.35) +
  geom_point(aes(colour=lake))+ scale_color_brewer(palette="Dark2")+
  ggtitle('E. puntang Head - Lake {closest_state}')+
  transition_states(lake,
                    transition_length = 2,
                    state_length = 3.5) +
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),
        legend.position = "bottom")
plot_bepun2
plot_hepun2
```

### *Acentrogobius janthinopterus*

```{r}
#Body
plot_bajan2 <-ggplot(data=coords_bajan,aes(x=x,y=y))+ 
  theme_scientific()+ ylim(-0.3,0.3) +
  geom_point(aes(colour=lake))+ scale_color_brewer(palette="Dark2")+
  ggtitle('A. janthinopterus body - Lake {closest_state}')+
  transition_states(lake,
                    transition_length = 2,
                    state_length = 3.5) +
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),
        legend.position = "bottom")
#Head
plot_hajan2 <-ggplot(data=coords_hajan,aes(x=x,y=y))+ 
  theme_scientific()+ ylim(-0.5,0.35) +
  geom_point(aes(colour=lake))+ scale_color_brewer(palette="Dark2")+
  ggtitle('A. janthinopterus Head - Lake {closest_state}')+
  transition_states(lake,
                    transition_length = 2,
                    state_length = 3.5) +
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),
        legend.position = "bottom")
plot_bajan2
plot_hajan2
```

### Saving .gif files

```{r}
anim_save("figures/supplements/epuntang_body.gif", plot_bepun2)
anim_save("figures/supplements/epuntang_head.gif", plot_hepun2)
anim_save("figures/supplements/ajanthinopterus_body.gif", plot_bajan2)
anim_save("figures/supplements/ajanthinopterus_head.gif", plot_hajan2)
```

Individual plots per location / Contrast between low and medium connectivity lakes

```{r}
#Body

mis01_bepun <- coords_bepun %>% 
  dplyr::filter(grepl('MIS01', lake))

plotb_mis01 <- ggplot() + theme_scientific() + ylim(-0.25,0.25) + xlim(-0.5, 0.7)+
  geom_point(data=mis01_bepun,
             aes(x=x,y=y)) +
  geom_point(data=con_bepun, #consensus shape of all locations
             aes(x=X,y=Y), size = 3, color="deepskyblue")+
  ggtitle("E. puntang - MIS01 (medium conn)")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank())

way04_bepun <- coords_bepun %>% 
  dplyr::filter(grepl('WAY04', lake))

plotb_way04 <- ggplot() + theme_scientific() + ylim(-0.25,0.25) + xlim(-0.5, 0.7)+
  geom_point(data=way04_bepun,
             aes(x=x,y=y)) +
  geom_point(data=con_bepun, #consensus shape of all locations
             aes(x=X,y=Y), size = 3, color="deepskyblue")+
  ggtitle("E. puntang - WAY04 (low conn)")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank())

plot_body <- plotb_mis01 + plotb_way04
plot_body

#Head

mis01_hepun <- coords_hepun %>% 
  dplyr::filter(grepl('MIS01', lake))

ploth_mis01 <- ggplot() + theme_scientific() + ylim(-0.4,0.3) + 
  geom_point(data=mis01_hepun,
             aes(x=x,y=y)) +
  geom_point(data=con_hepun,#consensus shape of all locations
             aes(x=X,y=Y), size = 3, color="deepskyblue")+
  ggtitle("E. puntang - MIS01 (medium conn)")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank())

way04_hepun <- coords_hepun %>% 
  dplyr::filter(grepl('WAY04', lake))

ploth_way04 <- ggplot() + theme_scientific() + ylim(-0.4,0.3) +
  geom_point(data=way04_hepun,
             aes(x=x,y=y)) +
  geom_point(data=con_hepun, #consensus shape of all locations
             aes(x=X,y=Y), size = 3, color="deepskyblue")+
  ggtitle("E. puntang - WAY04 (low conn)")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank())

plot_head <- ploth_mis01 + ploth_way04
plot_head

ggsave(plot_body,"figures/supplements/epun_body_connectivity.pdf")
ggsave(plot_head,"figures/supplements/epun_head_connectivity.pdf")
```


Combined

```{r}

body_comb <- coords_bepun %>% 
  dplyr::filter(grepl('MIS01|WAY04', lake))


p1 <- ggplot(data=body_comb,aes(x=x,y=y))+ 
  theme_scientific()+ ylim(-0.3,0.3) +
  geom_point(aes(colour=lake))+ scale_color_brewer(palette="Dark2")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),
        legend.position = "bottom")
p1


head_comb <- coords_hepun %>% 
  dplyr::filter(grepl('MIS01|WAY04', lake))


p2 <- ggplot(data=head_comb,aes(x=x,y=y))+ 
  theme_scientific()+ ylim(-0.4,0.4) +
  geom_point(aes(colour=lake))+ scale_color_brewer(palette="Dark2")+
  theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),
        legend.position = "bottom")
p2

p1 + p2 + plot_layout(guides = "collect") &
  theme(legend.position='bottom')
```

