---
title: "Buccal apparatus analyses"
author: "Stephanie Martinez"
date: "1/3/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We took *in situ* measurements on *E. puntang* individuals while we where on the field. These measurements included the protrusion, width and height of the mouth. These characteristics are related to feeding capabilities.

### Packages and data

```{r}

library(dplyr)
library(ggplot2)
library(patchwork)

mouth <- read.csv("data/mouthDim.csv", dec=".") 

mouth2 <- mouth %>% 
  filter(sp!="sp2") %>% 
  dplyr::select(mheight,mwidth,prot)
lakes <- mouth %>% 
  filter(sp!="sp2") %>%
  dplyr::select(location)
ID <- mouth %>% 
   filter(sp!="sp2") %>% 
   dplyr::select(ID)
```

### Correcting for size

Applying Mosimann's log-shape ratio

```{r}
var1_num <- as.numeric(ncol(mouth2))
size <- apply(mouth2, 1, prod)^(1/var1_num) 
log_mouth <- log(mouth2/ size)
LSR_mouth <- cbind(lakes,log_mouth)
```

### Distribution boxplots

```{r}
p1 <-ggplot(LSR_mouth, aes(x=location, 
                             y=mheight,
                             fill=location,alpha=0.1)) +
  scale_fill_manual(breaks=c("MIS01","MIS17","MIS20","WAR03","MIS14"), 
                    values = c("#ffa500","#33A02C","#474444","#df546c","#B2DF8A"))+
  geom_boxplot(show.legend = FALSE) + theme_scientific() + ggtitle("a)") +
  theme(plot.title = element_text(hjust = 0), legend.title = element_blank(),axis.text.x = element_blank())+
  labs(x="", y ="Mouth height (mm)")
  
p2 <- ggplot(LSR_mouth, aes(x=location, 
                    y=mwidth,
                    fill=location,alpha=0.1)) +
  scale_fill_manual(breaks=c("MIS01","MIS17","MIS20","WAR03","MIS14"), 
                    values = c("#ffa500","#33A02C","#474444","#df546c","#B2DF8A"))+
  geom_boxplot(show.legend = FALSE) + theme_scientific() + ggtitle("b)")+
  theme(plot.title = element_text(hjust = 0), legend.title = element_blank(), axis.text.x = element_blank())+
  labs(x="", y ="Mouth width (mm)")

p3<- ggplot(LSR_mouth, aes(x=location, 
                    y=prot,
                    fill=location,alpha=0.1)) +
  scale_fill_manual(breaks=c("MIS01","MIS17","MIS20","WAR03","MIS14"), 
                    values = c("#ffa500","#33A02C","#474444","#df546c","#8c008c"))+
  geom_boxplot(show.legend = FALSE) + theme_scientific() + ggtitle("c)")+
  theme(plot.title = element_text(hjust = 0), legend.title = element_blank())+
  labs(x="", y ="Protusion (mm)")

p1 / p2 / p3
boxplot_mouth <- p1 / p2 / p3
ggsave(boxplot_mouth, filename = "figures/supplements/boxplot_mouthD.pdf")

```

### Prinpical Component Analysis (PCA)

Preparing dataframe for pca

```{r}

mouth.pca <- cbind(ID, LSR_mouth)
df.pca <- mouth.pca %>% 
  remove_rownames(.) %>% 
  column_to_rownames(., var="ID") %>% 
  dplyr::select(mheight, mwidth, prot)
```

PCA

```{r}
pca <- prcomp(df.pca, scale. = TRUE)
```

PCA plot (ggplot) preparation

```{r}
pcaV <- as.data.frame(pca$x) %>% 
  dplyr::select(PC1, PC2) 
a1 <- summary(pca) 
axis <- as.data.frame(a1$importance) %>% 
  dplyr::select(PC1,PC2) %>% 
  mutate(PC1 = PC1*100, PC2 = PC2*100) 
```

```{r}
p1 <- ggplot()+ 
  theme_scientific()+
  geom_point(data=pcaV,  
             aes(x = PC1, 
                 y = PC2,
                 color =mouth.pca$location,
                 shape= mouth.pca$location), size = 3) +
  scale_color_manual(breaks=c("MIS01","MIS17","MIS20","WAR03","MIS14"), 
                    values = c("#ffa500","#33A02C","#474444","#df546c","#8c008c")) +
  scale_shape_manual(breaks=c("MIS01","MIS17","MIS20","WAR03","MIS14"),
                     values=c(15,16,1,4,17))+
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("PCA1 (",round(axis$PC1[2],2),"%)"), 
       y =paste("PCA2 (",round(axis$PC2[2],2),"%)")) +
  ggtitle("(a) PCA mouth dimensions ")+
  theme(plot.title = element_text(hjust = -0.15))

p2 <-fviz_pca_var(pca, col.circle="black", title="b) Variables", 
                 xlab="PCA1 (63.5%)",
                 ylab="PCA2 (23.5%)") + #Variables circles/eigenvalues
  xlim(c(-1.2,1.2)) +
  ylim(c(-1.2,1.2))

p3 <- p1+p2
ggsave(p3, filename = "figures/pca_mouthDimension.pdf")
```

ANOVA and pairwise test

```{r}

Anova(lm(pca$x[,1:2]~mouth.pca$location)) #taking away 1 dimension

pwc <- mouth.pca %>%
  gather(key = "variables", value = "value", mheight,mwidth,prot) %>%
  group_by(variables) %>%
  games_howell_test(value ~ location) %>%
  select(-estimate, -conf.low, -conf.high) %>% 
  as.data.frame() %>% 
  filter(p.adj.signif!="ns")

pwc
```
