---
title: "Geomorphometric Analysis"
author: "Stephanie Martinez"
date: '2020'
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Geomorphometric analysis

Here I present the code used in my geometric morphometric analysis. The morphometry (body and head views) of each species was analyzed separately. Thus, the markdown sections will be structure as such.

### Packages and data

Load .tps files that contains coordinate data, and a classifier dataframe containing environmental variables and location information (e.g. year, lake name, species (sp), temperature)

```{r packages, message=FALSE, warning=FALSE}
library("geomorph")
library("tidyr")
library("dplyr")
library("artyfarty")
library("ggpubr")
library("ggplot2")
```

**Abbreviations:**

-   epun = *Exyrias puntang* ; ajan = *Acentrogobius janthinopterus*

-   bepun = body *E. puntang ;* bajan = body *A. janthinopterus*

-   hepun = head *E. puntang ;* hajan = head A. janthinopterus

```{r data, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
body_epun <- readland.tps("data/body_epun.TPS", specID = "ID", warnmsg =T)
body_ajan <- readland.tps("data/body_ajan.TPS", specID = "ID", warnmsg =T)
head_epun <- readland.tps("data/head_epun.TPS", specID = "ID", warnmsg =T)
head_ajan <- readland.tps("data/head_ajan.TPS", specID = "ID", warnmsg =T)

class <- read.csv("data/classifiers.csv", header=T, dec=".", sep = ";") %>% 
  filter(sp!="nd") #remove nd (not identified) specimens

```

## *1- Exyrias puntang*

Adapt and clean the classifiers data frame to *E. puntang* .tps data structure

```{r epun clasifier}
class_epun <- class %>%
  filter(sp!="sp2") %>% 
  filter(!grepl('WAR04_G7_L1', ID)) %>% #outlier - removed
  droplevels()
```

### Superimposition of raw coordinates

```{r echo=TRUE, message=FALSE, warning=FALSE}
bepun_gpagen <- gpagen(body_epun, Proj=TRUE, print.progress = FALSE)
plot(bepun_gpagen)
title(main="Body E.puntang procrustes")
hepun_gpagen <- gpagen(head_epun, Proj=TRUE, print.progress = FALSE)
plot(hepun_gpagen)
title(main="Head E.puntang procrustes")

```

### PCA of superimposed coordinates

The PCA analysis was done using *geomorph*. Plotting was done by extracting values from the test and using *ggplot2.*

First, the PCA analysis

```{r pca analyzes}
pca_bepun <- gm.prcomp(bepun_gpagen$coords)
pca_hepun <- gm.prcomp(hepun_gpagen$coords)
```

Now, we can extract the values needed for plotting using *ggplot2.* We also set classifiers as factors for our plot.

```{r message=FALSE, warning=FALSE, results='hide'}

#Grouping factors 
lakes_epun <- as.factor(class_epun[,3])
conn_epun <- as.factor(class_epun[,6])
#extracting PCA component values for plot
comp.bepun <- as.data.frame(pca_bepun$x) %>% 
  dplyr::select(Comp1, Comp2) 
comp.hepun <- as.data.frame(pca_hepun$x) %>% 
  dplyr::select(Comp1, Comp2)
#extracting PCA variance for plot axis
a1 <- summary(pca_bepun) #make the summary an object to extract information
axis1 <- as.data.frame(a1$PC.summary) %>% 
  dplyr::select(Comp1,Comp2) %>% 
  mutate(Comp1 = Comp1*100, Comp2 = Comp2*100)
a2 <- summary(pca_hepun)  
axis2 <- as.data.frame(a2$PC.summary) %>% 
  dplyr::select(Comp1,Comp2) %>% 
  mutate(Comp1 = Comp1*100, Comp2 = Comp2*100)
```

Finally, we can produce a plot using *ggplot*

```{r pca plot ggplot}
#Body
plot_bepun <- ggplot()+ 
  theme_scientific()+
  geom_point(data=comp.bepun,  
             aes(x = Comp1, 
                 y = Comp2,
                 color =lakes_epun,
                 shape= lakes_epun), size = 3) +
  scale_color_manual(breaks=c("kar02","mis01","mis17","mis20","mis31","war03","war04","way04"), 
                     values = c("#000080","#ffa500","#33A02C","#474444","#1078bf","#df546c","#B2DF8A","#1be0e4")) +
  scale_shape_manual(breaks=c("kar02","mis01","mis17","mis20","mis31","war03","war04","way04"),
                     values=c(8,15,16,1,18,4,15,10))+
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("PCA1 (",round(axis1$Comp1[2],2),"%)"), 
       y =paste("PCA2 (",round(axis1$Comp2[2],2),"%)")) +
  ggtitle("(a) Body landmarks")+
  theme(plot.title = element_text(hjust = -0.15))
#Head
plot_hepun <- ggplot()+ 
  theme_scientific()+
  geom_point(data=comp.hepun,  
             aes(x = Comp1, 
                 y = Comp2,
                 color =lakes_epun,
                 shape= lakes_epun), size = 3) +
  scale_color_manual(breaks=c("kar02","mis01","mis17","mis20","mis31","war03","war04","way04"), 
                     values = c("#000080","#ffa500","#33A02C","#474444","#1078bf","#df546c","#B2DF8A","#1be0e4")) +
  scale_shape_manual(breaks=c("kar02","mis01","mis17","mis20","mis31","war03","war04","way04"),
                     values=c(8,15,16,1,18,4,15,10))+
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("PCA1 (",round(axis2$Comp1[2],2),"%)"), 
       y =paste("PCA2 (",round(axis2$Comp2[2],2),"%)")) +
  ggtitle("(b) Head landmarks")+
  theme(plot.title = element_text(hjust = -0.15))
#Plot
ggarrange(plot_bepun, plot_hepun, common.legend=TRUE, legend = "bottom")
```

To select and pick warp grids in the PCA plot, you can use the function *picknplot*.*shape* from the package *geomorph*. However, these warps can only be extracted from a plot that uses the *plot* function. On the console, you can choose whether to select another point in the plot, and if you wish to save it as a .png image. See example bellow.

```{r warp plots, eval=FALSE, message=FALSE, warning=FALSE}
warp_bepun <- plot(pca_bepun, main="Body PCA Exyrias puntang")
points(pca_bepun$x,pch=19, col=lakes_epun)
legend("topleft", levels(lakes_epun), pch=19, col=1:8)
picknplot.shape(warp_bepun) #select warp grid in PCA plot

```

## *2- Acentrogobius janthinopterus*

For this species, we followed the same process a describe before for *Exyrias puntang*

Adapt and clean the classifiers data frame to *E. puntang* .tps data structure

```{r ajan clasifier}
class_ajan <- class %>%
  filter(sp!="sp1") %>% 
  filter(!grepl('KAR02', ID)) %>% 
  droplevels()
```

### Superimposition of raw coordinates

```{r echo=TRUE, message=FALSE, warning=FALSE}
bajan_gpagen <- gpagen(body_ajan, Proj=TRUE, print.progress = FALSE)
plot(bajan_gpagen)
title(main="Body A.janthinopterus procrustes")
hajan_gpagen <- gpagen(head_ajan, Proj=TRUE, print.progress = FALSE)
plot(hajan_gpagen)
title(main="Head A. janthinopterus procrustes")

```

### PCA of superimposed coordinates

PCA analysis

```{r pca ajan analyzes}
pca_bajan <- gm.prcomp(bajan_gpagen$coords)
pca_hajan <- gm.prcomp(hajan_gpagen$coords)
```

Extracting values needed for plotting using *ggplot2.* We also set classifiers as factors for our plot.

```{r message=FALSE, warning=FALSE, results='hide'}

#Grouping factors 
lakes_ajan <- as.factor(class_ajan[,3])
conn_ajan <- as.factor(class_ajan[,6])
#extracting PCA component values for plot
comp.bajan <- as.data.frame(pca_bajan$x) %>% 
  dplyr::select(Comp1, Comp2) 
comp.hajan <- as.data.frame(pca_hajan$x) %>% 
  dplyr::select(Comp1, Comp2)
#extracting PCA variance for plot axis
a3 <- summary(pca_bajan) #make the summary an object to extract information
axis3 <- as.data.frame(a3$PC.summary) %>% 
  dplyr::select(Comp1,Comp2) %>% 
  mutate(Comp1 = Comp1*100, Comp2 = Comp2*100)
a4 <- summary(pca_hajan)  
axis4 <- as.data.frame(a4$PC.summary) %>% 
  dplyr::select(Comp1,Comp2) %>% 
  mutate(Comp1 = Comp1*100, Comp2 = Comp2*100)
```

Finally, we can produce a plot using *ggplot* for *Acentrogobius janthinopterus*

```{r pca plot ggplot ajan}
#Body
plot_bajan <- ggplot()+ 
  theme_scientific()+
  geom_point(data=comp.bajan,  
             aes(x = Comp1, 
                 y = Comp2,
                 color =lakes_ajan,
                 shape= lakes_ajan), size = 3) +
  scale_color_manual(breaks=c("mis14","mis31","raj04","war01","war03","war04"), 
                     values=c("#8c008c","#1078bf","#ff66b3","#000000","#fd1530","#B2DF8A")) +
  scale_shape_manual(breaks=c("mis14","mis31","raj04","war01","war03","war04"),
                     values=c(17,18,16,0,4,15))+
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("PCA1 (",round(axis3$Comp1[2],2),"%)"), 
       y =paste("PCA2 (",round(axis3$Comp2[2],2),"%)"))+
  theme(plot.title = element_text(hjust = -0.15)) +
  ggtitle("(a) Head Landmarks")
#Head
plot_hajan <- ggplot()+ 
  theme_scientific()+
  geom_point(data=comp.hajan,  
             aes(x = Comp1, 
                 y = Comp2,
                 color =lakes_ajan,
                 shape= lakes_ajan), size = 3) +
  scale_color_manual(breaks=c("mis14","mis31","raj04","war01","war03","war04"), 
                     values=c("#8c008c","#1078bf","#ff66b3","#000000","#fd1530","#B2DF8A")) +
  scale_shape_manual(breaks=c("mis14","mis31","raj04","war01","war03","war04"),
                     values=c(17,18,16,0,4,15))+
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("PCA1 (",round(axis4$Comp1[2],2),"%)"), 
       y =paste("PCA2 (",round(axis4$Comp2[2],2),"%)"))+
  ggtitle("(a) Head Landmarks")+
  theme(plot.title = element_text(hjust = -0.15))
#Plot
ggarrange(plot_bajan, plot_hajan, common.legend=TRUE, legend = "bottom")
```

## 3- Statistical analyses

### MANOVA

To test if there were differences in the morphology of these species across lakes, a MANOVA test was performed using *ProcD.lm*

First, the superimposed coordinate data has to be arrange as a data frame

```{r manova dataframe}
#Exyrias puntang
gdf_bepun <- geomorph.data.frame(bepun_gpagen, 
                                lake = class_epun$lake, 
                                conn = class_epun$connec,
                                temp = class_epun$temp)
                                
gdf_hepun <- geomorph.data.frame(hepun_gpagen, 
                                lake = class_epun$lake, 
                                conn = class_epun$connec,
                                temp = class_epun$temp)

# Acentrogobius janthinopterus
gdf_bajan <- geomorph.data.frame(bajan_gpagen, 
                               lake = class_ajan$lake, 
                               conn = class_ajan$connec,
                               temp = class_ajan$temp)
gdf_hajan <- geomorph.data.frame(bajan_gpagen, 
                               lake = class_ajan$lake, 
                               conn = class_ajan$connec,
                               temp = class_ajan$temp)

```

MANOVAs using each lake and connectivity levels as grouping factors.

**Note:** RRPP uses randomize raw values

```{r manova, results='hide'}
# Exyrias puntang
fit1_bepun <- procD.lm(bepun_gpagen$coords ~ lake, 
                      data = gdf_bepun, iter = 999, 
                      RRPP = FALSE, print.progress = FALSE)
fit2_bepun <- procD.lm(bepun_gpagen$coords ~ conn, 
                     data = gdf_bepun, iter = 999, 
                     RRPP = FALSE, print.progress = FALSE)
fit3_hepun <- procD.lm(hepun_gpagen$coords ~ lake, 
                      data = gdf_hepun, iter = 999, 
                      RRPP = FALSE, print.progress = FALSE)
fit4_hepun <- procD.lm(hepun_gpagen$coords ~ conn, 
                     data = gdf_hepun, iter = 999, 
                     RRPP = FALSE, print.progress = FALSE)
summary(fit1_bepun)
summary(fit2_bepun)
summary(fit3_hepun)
summary(fit4_hepun)

# Acentrogobius janthinopterus
fit1_bajan <- procD.lm(bajan_gpagen$coords ~ lake, 
                     data = gdf_bajan, iter = 999, 
                     RRPP = FALSE, print.progress = FALSE)
fit2_bajan <- procD.lm(bajan_gpagen$coords ~ conn, 
                     data = gdf_bajan, iter = 999, 
                     RRPP = FALSE, print.progress = FALSE)
fit3_hajan <- procD.lm(hajan_gpagen$coords ~ lake, 
                     data = gdf_hajan, iter = 999, 
                     RRPP = FALSE, print.progress = FALSE)
fit4_hajan <- procD.lm(hajan_gpagen$coords ~ conn, 
                     data = gdf_hajan, iter = 999, 
                     RRPP = FALSE, print.progress = FALSE)
summary(fit1_bajan)
summary(fit2_bajan)
summary(fit3_hajan)
summary(fit4_hajan)
```

### Post-hoc Pairwise test

To see which lakes where contributing to the differences in morphology, we performed a post-hoc pairwise test from the RRP package.

The test uses *"dist"* which is the distance between LS means. By specifying a confidence level, we are given upper confidence limits (UCL) from the distributions of pairwise distances.

If the observed distances are larger than the UCL, the p-value is less than 1 - confidence (0.005), the distances are "significantly" different.

```{r pairwise tests, results='hide'}
# Exyrias puntang
PW_bepun <- pairwise(fit1_bepun, groups = gdf_bepun$lake)
summary(PW_bepun, test.type = "dist", confidence = 0.95, stat.table = FALSE)
PW_hepun <- pairwise(fit3_hepun, groups = gdf_hepun$lake)
summary(PW_hepun, test.type = "dist", confidence = 0.95, stat.table = FALSE)
# Acentrogobius janthinopterus
PW_bajan <- pairwise(fit1_bajan, groups = gdf_bajan$lake)
summary(PW_bepun, test.type = "dist", confidence = 0.95, stat.table = FALSE)
PW_hajan <- pairwise(fit3_hajan, groups = gdf_hajan$lake)
summary(PW_hepun, test.type = "dist", confidence = 0.95, stat.table = FALSE)
```
