---
title: "Linear Morphometric Analyses"
author: "Stephanie Martinez"
date: "2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For the linear morphometric analyses, we used the interlandmark distances obtained from the previous script "analyses/landmark_distances_calculation.Rmd"

### Packages and data

```{r packages and data, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(factoextra)
library(GGally)
library(candisc)
library(ggforce)
library(patchwork)
library(artyfarty)
library(rstatix)

dist_epun <- read.csv("data/distances_epun.csv",dec=".") %>% 
  slice(-9) ##removing G5 because it contains a zero value
dist_ajan <- read.csv("data/distances_ajan.csv",dec=".")
```

### Size correction

To correct for the size differences among individuals we used the log-shape ratio approach by Mosimann (1970), and followed the script and considerations by Julien (2013).

First, we need to change the dataframe formats required for performing PCA analysis (further on)

```{r}
dist2_epun <- dist_epun %>% 
  dplyr::select(-X,-lake,-goby) %>% 
  remove_rownames(.) %>% 
  column_to_rownames(., var="ID") 
dist2_ajan <- dist_ajan %>% 
  dplyr::select(-X,-lake,-goby) %>% 
  remove_rownames(.) %>% 
  column_to_rownames(., var="ID") 
```

Calculating the geometric mean to obtain a "size" variable

```{r}
#E. puntang
var_epun <- as.numeric(ncol(dist2_epun))
size_epun <- apply(dist2_epun, 1, prod)^(1/var_epun) 
# A. janthinopterus
var_ajan <- as.numeric(ncol(dist2_ajan))
size_ajan <- apply(dist2_ajan, 1, prod)^(1/var_ajan) 
```

Calculating the log-shape ratios or Mosimann's shape vectors and produce a new dataframe called LSR

```{r}
LSR_epun <- log(dist2_epun/ size_epun)
LSR_ajan <- log(dist2_ajan/ size_ajan)
```

### Correlation analysis

To avoid colinearity between variables, we did an correlation test. Furthermore, we discarded in further analysis the variables that presented more than 75% (\<0.75) correlation.

For this, we use ggpairs to visualize the correlation between all the variables. Since in these cases we have more than 20 variables, we'll separate the graphs for better visualization

```{r message=FALSE, warning=FALSE, results='hold'}
#E. puntang
corr1_epun <- ggpairs(LSR_epun, columns = 1:15, ggplot2::aes(colour=dist2_epun$lake))
corr2_epun <-ggpairs(LSR_epun, columns = 16:29, ggplot2::aes(colour=dist2_epun$lake))
corr1_epun
corr2_epun
# A. janthinopterus
corr1_ajan <- ggpairs(LSR_ajan, columns = 1:11, ggplot2::aes(colour=dist2_ajan$lake))
corr2_ajan <-ggpairs(LSR_ajan, columns = 12:24, ggplot2::aes(colour=dist2_ajan$lake))
corr1_ajan
corr2_ajan
#Saving images
ggsave(corr1_epun,filename = "figures/supplements/corr1_epun.pdf")
ggsave(corr2_epun,filename = "figures/supplements/corr2_epun.pdf")
ggsave(corr1_ajan,filename = "figures/supplements/corr1_ajan.pdf")
ggsave(corr2_ajan,filename = "figures/supplements/corr2_ajan.pdf")
```

Based on these results we took out the following variables (interlandmark distances) for each species:

***Exyrias puntang:***

-   TL, HP, PH, PH2, PEH, PEH2, PRP, POP, PRF, DD, CPD, PAL, ABL

***Acentrogobius janthinopterus:***

-   TL, HP, PH, PH2, PRP, POP, PRF, DD, CPD, PAL, ABL

```{r}
#E. puntang
corr_epun <- LSR_epun %>% 
  dplyr::select(-TL,-HP,-PH,-PH2,-PEH,-PEH2,-PRP,-POP,-PRF,-DD,-CPD,-PAL,-ABL)
# write.csv(corr_epun,"data/dist_corr_epun.csv")

#A. janthinopterus
corr_ajan <- LSR_ajan %>% 
  dplyr::select(-TL,-HP,-PH,-PH2,-PRP,-POP,-PRF,-DD,-CPD,-PAL,-ABL)
# write.csv(corr_ajan,"data/dist_corr_ajan.csv")
```

### PCA analysis

```{r message=FALSE, warning=FALSE}
#E. puntang
pca_epun <- prcomp(corr_epun, scale. = TRUE)
Anova(lm(pca_epun$x[,1:3]~dist_epun$lake*size_epun)) #included size to see if there was an allometric effect
#A. janthinopterus
pca_ajan <- prcomp(corr_ajan, scale. = TRUE)
Anova(lm(pca_ajan$x[,1:3]~dist_ajan$lake*size_ajan))
```

For plotting the PCA analysis, there are simple and easy to use functions such as "fviz_pca_ind". However, since we had so many groups (lakes) and we had to visualize both species, we manually created the plots in ggplot2 to have more customization options.

**Extracting data from prcomp**

```{r}
#E. puntang
pcaV.epun <- as.data.frame(pca_epun$x) %>% 
  dplyr::select(PC1, PC2,PC3) #extracting PCA values for each specimen (dataframe)
a1 <- summary(pca_epun) #extracting PCA variance for plot axis 
axis_epun <- as.data.frame(a1$importance) %>% 
  dplyr::select(PC1,PC2,PC3) %>% 
  mutate(PC1 = PC1*100, PC2 = PC2*100, PC3=PC3*100)  #extracting PCA variance for plot axis and turn it into percentage
#A. janthinopterus
pcaV.ajan <- as.data.frame(pca_ajan$x) %>% 
  dplyr::select(PC1, PC2,PC3) 
a2 <- summary(pca_ajan) 
axis_ajan <- as.data.frame(a2$importance) %>% 
  dplyr::select(PC1,PC2,PC3) %>% 
  mutate(PC1 = PC1*100, PC2 = PC2*100, PC3=PC3*100)
```

### **Plots**

**Adding classifiers**

```{r}
lakes_epun <- as.factor(dist_epun[,3])
lakes_ajan <- as.factor(dist_ajan[,3])
```

**PCA Plots**

```{r message=FALSE, warning=FALSE, results='hold'}
#E. puntang
p1_epun <- ggplot()+ 
  theme_scientific()+
  geom_point(data=pcaV.epun,  
             aes(x = PC1, 
                 y = PC2,
                 color =lakes_epun,
                 shape= lakes_epun), size = 3) +
  scale_color_manual(breaks=c("KAR02","MIS01","MIS17","MIS20","MIS31","WAR03","WAR04","WAY04"), 
                     values = c("#000080","#ffa500","#33A02C","#474444","#1078bf","#df546c","#B2DF8A","#1be0e4")) +
  scale_shape_manual(breaks=c("KAR02","MIS01","MIS17","MIS20","MIS31","WAR03","WAR04","WAY04"),
                     values=c(8,15,16,1,18,4,15,10))+
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("PCA1 (",round(axis_epun$PC1[2],2),"%)"), 
       y =paste("PCA2 (",round(axis_epun$PC2[2],2),"%)")) +
  ggtitle("(a) PCA E.puntang")+
  theme(plot.title = element_text(hjust = -0.15))

p2_epun <-fviz_pca_var(pca_epun, col.circle="black", title="b) Variables", 
             xlab="PCA1 (41.2%)",
             ylab="PCA2 (24.5%)") + #Variables circles/eigenvalues
  xlim(c(-1.2,1.2)) +
  ylim(c(-1.2,1.2))

p1_epun + p2_epun
p3_epun <- p1_epun + p2_epun

# A. janthinopterus
p1_ajan <- ggplot()+ 
  theme_scientific()+
  geom_point(data=pcaV.ajan,  
             aes(x = PC1, 
                 y = PC2,
                 color =lakes_ajan,
                 shape= lakes_ajan), size = 3) +
  scale_color_manual(breaks=c("MIS14","MIS31","RAJ04","WAR01","WAR03","WAR04"), 
                     values=c("#8c008c","#1078bf","#ff66b3","#000000","#fd1530","#B2DF8A")) +
  scale_shape_manual(breaks=c("MIS14","MIS31","RAJ04","WAR01","WAR03","WAR04"),
                     values=c(17,18,16,0,4,15))+
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("PCA1 (",round(axis_ajan$PC1[2],2),"%)"), 
       y =paste("PCA2 (",round(axis_ajan$PC2[2],2),"%)"))+
  theme(plot.title = element_text(hjust = -0.15)) +
  ggtitle("(a) PCA A. janthinopterus")

p2_ajan <- fviz_pca_var(pca_ajan, col.circle="black", title="b) Variables", 
                 xlab="PCA1 (64.28%)",
                 ylab="PCA2 (8.51%)") + #Variables circles/eigenvalues
  xlim(c(-1.2,1.2)) +
  ylim(c(-1.2,1.2))

p1_ajan + p2_ajan
p3_ajan <- p1_ajan + p2_ajan

#save plots
ggsave(p3_epun, filename = "figures/supplements/pca_distances_epun.pdf")
ggsave(p3_ajan, filename = "figures/supplements/pca_distances_ajan.pdf")
```

### Canonical discriminate analysis

To assess which variables (interlandmark distances) provides the maximal separation and differentiation between the lakes, a canonical discriminant analysis was done.

First, we have to do a linear model and then apply the candisc funtion

```{r lm}
#E. puntang
epun_lm <- lm(cbind(ED, BD, MBL, OW, PRO, VF, PFB, PCL, ML, LJ)~dist_epun$lake, data=corr_epun)
#A. janthinopterus 
ajan_lm <- lm(cbind(ADL,ML,PRO,LJ,PCL,POD,ED,MBL,PFB,
                          BD,POO)~dist_ajan$lake, data=corr_ajan)
```

The candisc analysis can be represented in one or two dimensions. Both ways can help interpret the data and how the variables are interacting.

```{r}
#E. puntang
epun_can2d <- candisc(epun_lm, data=corr_epun)
epun_can1d <- candisc(epun_lm, data=corr_epun, ndim=1)
#A. janthinopterus
ajan_can2d <- candisc(ajan_lm, data=corr_ajan)
ajan_can1d <- candisc(ajan_lm, data=corr_ajan, ndim=1)

```

You can plot the results by using the basic plot function on the candisc object

```{r}
plot(epun_can2d)
plot(epun_can1d)

```

However, for better visualization we preferred using ggplot. As before (PCA plots) we need to extract certain elements from the candisc test and place them as datagrames.

**Extracting vector loadings (structure) and scores**

```{r}
#E. puntang
str.epun <- epun_can2d$structure %>% 
  as.data.frame() %>% 
  dplyr::select(-Can3) %>% 
  rownames_to_column(., var="Var") 
scores.epun<- epun_can2d$scores %>% 
  as.data.frame() %>% 
  rename(lake = 1)
#A. janthinopterus
str.ajan <- ajan_can2d$structure %>% 
  as.data.frame() %>% 
  dplyr::select(-Can3) %>% 
  rownames_to_column(., var="Var") 
scores.ajan<- ajan_can2d$scores %>% 
  as.data.frame() %>% 
  rename(lake = 1)
```

**2-Dimension CANDISC plots**

*E. puntang*

```{r}
p_epun_can2d <- ggplot()+ 
  theme_scientific()+ 
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("Can1 (",round(epun_can2d$pct[1],1),"%)"), 
       y =paste("Can2 (",round(epun_can2d$pct[2],1),"%)"))+
  geom_point(data=scores.epun, 
             aes(x = Can1, 
                 y = Can2, 
                 color = lake,
                 shape = lake), size=3) + 
  scale_color_manual(breaks=c("KAR02","MIS01","MIS17","MIS20","MIS31","WAR03","WAR04","WAY04"), 
                     values = c("#000080","#ffa500","#33A02C","#474444","#1078bf","#df546c","#B2DF8A","#1be0e4")) +
  scale_shape_manual(breaks=c("KAR02","MIS01","MIS17","MIS20","MIS31","WAR03","WAR04","WAY04"),
                     values=c(8,15,16,1,18,4,15,10)) +
  
  ## Adding Vectors
  geom_segment(aes(x=0,y=0,
                   xend=8*str.epun$Can1, 
                   yend=8*str.epun$Can2),
               color="black",alpha=.5,arrow=arrow(length=unit(0.3,"cm"))) +
  ## Adding Vectors label
  annotate("text", label = paste(str.epun$Var),
           fontface = 'italic',
           x = 9.5*str.epun$Can1, 
           y =8.3*str.epun$Can2,
           color="black",alpha=.5)+
  # Adding group mean
  geom_point(data=epun_can2d$means, 
             aes(x = Can1, 
                 y = Can2,
                 color=unique(scores.epun$lake)),
             shape=3,size=3, stroke=2, show.legend = FALSE)
p_epun_can2d
ggsave(p_epun_can2d, filename = "figures/candisc2d_epun.pdf")
```

*A. janthinopterus*

```{r}
p_ajan_can2d <- ggplot()+ 
  theme_scientific()+ 
  theme(legend.position = "bottom", legend.title = element_blank())+
  labs(x=paste("Can1 (",round(ajan_can2d$pct[1],1),"%)"), 
       y =paste("Can2 (",round(ajan_can2d$pct[2],1),"%)"))+
  geom_point(data=scores.ajan, 
             aes(x = Can1, 
                 y = Can2, 
                 color = lake,
                 shape = lake), size=3) + 
  scale_color_manual(breaks=c("MIS14","MIS31","RAJ04","WAR01","WAR03","WAR04"), 
                     values=c("#8c008c","#1078bf","#ff66b3","#000000","#fd1530","#B2DF8A")) +
  scale_shape_manual(breaks=c("MIS14","MIS31","RAJ04","WAR01","WAR03","WAR04"),
                     values=c(17,18,16,0,4,15)) +
  
  # Adding Vectors
  geom_segment(aes(x=0,y=0,
                  xend=8*str.ajan$Can1, 
                  yend=8*str.ajan$Can2),
               color="black",alpha=.5,arrow=arrow(length=unit(0.3,"cm"))) +
  # Adding Vectors label
  annotate("text", label = paste(str.ajan$Var),
                   fontface = 'italic',
                   x = 9.5*str.ajan$Can1, 
                  y =8.3*str.ajan$Can2,
                   color="black",alpha=.5)+
  # Adding group mean
  geom_point(data=ajan_can2d$means, 
             aes(x = Can1, 
                 y = Can2,
                 color=unique(scores.ajan$lake)),
             shape=3,size=3, stroke=2, show.legend = FALSE)
p_ajan_can2d
ggsave(p_ajan_can2d, filename = "figures/candisc2d_ajan.pdf")
```

**1-Dimension CANDISC plots**

As a complement, we can also observe the one dimensional visualization of the discriminate analysis next to the loading score of each discriminant (interlandmark distance)

*E. puntang*

```{r}
#Canonical scores

p1_epun_can1d <- ggplot(scores.epun, aes(x=lake, 
                             y=Can1,
                             fill=lake,alpha=0.1)) +
  scale_fill_manual(breaks=c("KAR02","MIS01","MIS17","MIS20","MIS31","WAR03","WAR04","WAY04"), 
                     values = c("#000080","#ffa500","#33A02C","#474444","#1078bf","#df546c","#B2DF8A","#1be0e4")) +
  geom_boxplot(show.legend = FALSE) + theme_scientific() +
  ggtitle("a.Canonical scores")+
  theme(plot.title = element_text(hjust = 0),
        axis.text.x = element_text(angle = 90))+
  labs(x="", y =paste("Can1 (",round(epun_can1d$pct[1],2),"%)"))
p1_epun_can1d

#Structure scores / discriminant loadings

structure_epun <- as.data.frame(epun_can1d$structure)
structure_epun <- structure_epun %>% 
  rownames_to_column(.,var="Vector") %>% 
  rowid_to_column(var = "ID") %>% 
  mutate(Can1_P = 100*Can1)


p2_epun_can1d <- ggplot(structure_epun) + theme_scientific()+
  geom_segment(aes(x=reorder(Vector,-Can1_P), xend=Vector, y=0, yend=Can1_P),
               arrow=arrow(length=unit(0.3,"cm")),color="blue")+
  labs(x="", y =paste("Loadings (%)")) +
  ggtitle("b.Discriminant loadings") +
  theme(plot.title = element_text(hjust = 0),
        axis.text.x = element_text(angle = 90))

p2_epun_can1d
p_epun_can1d <- p2_epun_can1d + p2_epun_can1d
ggsave(p_epun_can1d, filename = "figures/supplements/candisc1d_epun.pdf")
```

*A. janthinopterus*

```{r}
p1_ajan_can1d  <- ggplot(scores.ajan, aes(x=lake, 
                                y=Can1,
                                fill=lake,alpha=0.1)) +
  scale_fill_manual(breaks=c("MIS14","MIS31","RAJ04","WAR01","WAR03","WAR04"), 
                     values=c("#8c008c","#1078bf","#ff66b3","#000000","#fd1530","#B2DF8A")) +
  geom_boxplot(show.legend = FALSE) + theme_scientific() +
  ggtitle("a.Canonical scores")+
  theme(plot.title = element_text(hjust = 0),
        axis.text.x = element_text(angle = 90),
        legend.position = "bottom", legend.title = element_blank())+
  labs(x="", y =paste("Can1 (",round(ajan_can1d$pct[1],2),"%)"))
p1_ajan_can1d
  
#Structure scores / discriminant loadings

structure_ajan <- as.data.frame(ajan_can1d$structure)
structure_ajan <- structure_ajan %>% 
  rownames_to_column(.,var="Vector") %>% 
  rowid_to_column(var = "ID") %>% 
  mutate(Can1_P = 100*Can1)
  

p2_ajan_can1d <- ggplot(structure_ajan) + theme_scientific()+
  geom_segment(aes(x=reorder(Vector,-Can1_P), xend=Vector, y=0, yend=Can1_P),
               arrow=arrow(length=unit(0.3,"cm")),color="blue")+
  labs(x="", y =paste("Loadings (%)")) +
  ggtitle("b.Discriminant loadings") +
  theme(plot.title = element_text(hjust = 0),
        axis.text.x = element_text(angle = 90))

p2_ajan_can1d
p_ajan_can1d <- p1_ajan_can1d + p2_ajan_can1d
ggsave(p_ajan_can2d, filename = "figures/supplements/candisc1d_ajan.pdf")

```

### Statistical tests

To determine if the discrimination among lakes was significant we performed a MANOVA based on the linear model done before

```{r}
#E. puntang
epun_mod <- cbind(lakes_epun, dist_epun) %>% 
  dplyr::select(-X)

mod.epun <- lm(cbind(PFB,MBL,BD,PCL,PRO,ED,ML,LJ,OW,VF)~lake,
               data=epun_mod)
Manova(mod.epun, test="Pillai")

#A. janthinopterus
ajan_mod <- cbind(lakes_ajan, dist_ajan)
  
mod.ajan <- lm(cbind(ADL,ML,PRO,LJ,PCL,POD,ED,MBL,PFB,BD,POO)~lake, 
               data=ajan_mod)
Manova(mod.ajan, test="Pillai")

```

To see which interactions between variables are causing these differences, we can performe a pairwise test.

```{r}
#E. puntang
pwc_epun <- epun_mod %>%  
  gather(key = "variables", value = "value",ED,HL,BD,MBL,OW,PRO,VF,PFB,PCL,ML,LJ) %>%
  group_by(variables) %>%
  games_howell_test(value ~ lake) %>%
  select(-estimate, -conf.low, -conf.high) %>% 
  as.data.frame() %>% 
  filter(p.adj.signif!="ns")
pwc_epun
# A. janthinopterus
pwc_ajan <- ajan_mod %>%  
  gather(key = "variables", value = "value",ADL,ML,PRO,LJ,PCL,POD,ED,MBL,PFB,BD,POO) %>%
  group_by(variables) %>%
  games_howell_test(value ~ lake) %>%
  select(-estimate, -conf.low, -conf.high) %>% 
  as.data.frame() %>% 
  filter(p.adj.signif!="ns")
pwc_epun
```
