---
title: "TPS_reshape"
author: "Stephanie Martinez"
date: "2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reshaping .TPS files into .cvs files

The following markdown will show the steps to reshaping a .TPS files into .cvs files that were used in the linear morphometric analyses. This process was repeated for each fish view (body, head, and ventral), and for each species: *Exyrias puntang (*epun) and *Acentrogobius janthinopterus* (ajan)\
\
**Note**: The names of each sample (image) contains the necessary information to identify sample number, location, and view. This is necessary for the later steps of the reshaping process

**Note:** ventral view measures were only taken for *Exyrias puntgang* individuals

### Packages

```{r packages, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(geomorph)
library(reshape2)
```

The number of landmarks (landm) used on the files should be replaced accordingly:

-   Body view = 18

-   Head view = 12

-   Ventral view = 5

**Note:** "epun" stands for *Exyrias puntang*

```{r data, echo=FALSE}
coords <- readland.tps("data/body_epun.TPS", specID ="ID", warnmsg = T)
landm<-18
coords_2d <- as.data.frame(two.d.array(coords)) # put in a 2D array with species as the x input and coordinates (x&y) as y input

```

### Reshaping 2d array

```{r, echo=TRUE, message=FALSE, warning=FALSE}
var1 <- as.data.frame(matrix(c("x","y"), landm, 2, byrow=T))
var1$x <- paste(var1$V1, rownames(var1), sep="")
var1$y <- paste(var1$V2, rownames(var1), sep="")
var2 <- as.vector(t(var1[,-1:-2]))  # vector containing names of the variables
names(coords_2d) <- var2 # dataframe with correct variable names
coords1 <- coords_2d
coords1$ID <- rownames(coords_2d)
coords2 <- melt(coords1, id.vars="ID")
coords2$axis <- substr(coords2$variable,1,1)
coords2$point <- as.numeric(substr(coords2$variable,2,4))
coords_final <- dcast(coords2, ID + point~axis) 
```

```{r}
head(coords_final)
```

### Adjusting column names and format

So far, each landmark is identified with a number code. In order to later calculate inter-landmark distances using *r2d2,* it is necessary to assign a unique letter to each landmark (column pointL)\
\
In this case, for each view (body, head, ventral) a different letter range is required:

-   Body = [1:18]

-   Head = [1:12]

-   Ventral = [1:5]

Moreover, with *separate* with can create new columns based on the sample information that ID provides

```{r message=FALSE, warning=FALSE}
coords_final2 <- coords_final %>% 
  separate(ID, into = c("lake", "goby","side"), sep = "_", remove = FALSE) %>%
  dplyr::select(-side) %>% 
  mutate(pointL = rep(LETTERS[1:18],length(unique(.$ID)))) %>% #change letter range accordingly
  relocate(pointL, .after=point) 
head(coords_final2)

```

### Export as a csv

Remember to change the output file name according to each view

```{r}

write.csv(coords_final2,'data/coords_body_epun.csv')
```

The same process was done for each view and species obtaining as an output the following dataframes:

-   "data/coords_body_epun.csv"

-   "data/coords_head_epun.csv"

-   "data/coords_ventral_epun.csv"

-   "data/coords_body_ajan.csv"

-   "data/coords_head_ajan.csv"
